---
title: "Module 1 - Introducing Data"
subtitle: "edX EdinburghX Statistics"
author: "Clare Gibson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: cosmo
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This module looks at the techniques for describing, presenting and summarising data. Throughout this module we will make use of a data set on primary school performance in Surrey and Hampshire to illustrate some of the key concepts.

We will make use of the following libraries:
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```


## Loading the example data
Let's read in the data set we will use for this module and look at the first 6 rows of data. This data set describes Key Stage 2 performance for primary schools in Surrey for the 2018-2019 academic year. It was published by the UK government.
```{r read, message=FALSE, warning=FALSE}
performance <- read_csv("data-in/2018-2019_936_ks2final.csv") %>% 
  clean_names()

head(performance)
```

There are over 300 columns of data in this file. For the purposes of this module we can condense this down to a more manageable number of columns.
```{r select}
performance <- performance %>% 
  select(urn,
         schname,
         nftype,
         reldenom,
         totpups,
         matprog,
         writprog,
         readprog)

head(performance)
```

Looking at the data types that R has determined, we will need to make some adjustments.
```{r data types, message=FALSE, warning=FALSE}
performance <- performance %>% 
  mutate(across(totpups:readprog, as.numeric))

head(performance)
```

# What is data?
Data is information and comes in different types:

* **Qualitative** data describes qualities. It is descriptive. There are two types of qualitative data:
    + **Ordinal** tells you about ordering (e.g. it ranks 8 out of 10)
    + **Nominal** simply describes or categorises something (e.g. it's pink)
* **Quantitative** data talks about quantities or numbers (e.g. 15cm high). There are two types of quantitative data:
    + **Discrete** can only take a set of possible values (e.g. £1.90)
    + **Continuous** can take on any value in a range (e.g. 12, 12.31. 12.34562)

In our example data set we have the following fields and data types.

Column Name     Data Type               Description
--------------- ----------------------  -----------------
urn             Qualitative Nominal     Unique Reference Number of the school
schname         Qualitative Nominal     School name
nftype          Qualitative Nominal     School type
reldenom        Qualitative Nominal     Religious denomination
totpups         Quantitative Discrete   Total number of pupils
matprog         Quantitative Continuous Maths progress measure
writprog        Quantitative Continuous Writing progress measure
readprog        Quantitative Continuous Reading progress measure

Table: Fields and data types in the example data set