---
title: "Module 1 - Introducing Data"
subtitle: "edX EdinburghX Statistics"
author: "Clare Gibson"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This module looks at the techniques for describing, presenting and summarising data. Throughout this module we will make use of a data set on primary school performance in Surrey and Hampshire to illustrate some of the key concepts.

We will make use of the `tidyverse` and `janitor` packages:
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```

## Loading the example data
Let's read in the data set we will use for this module and look at the first 6 rows of data. This data set describes Key Stage 2 performance for primary schools in Surrey for the 2018-2019 academic year. It was published by the UK government.
```{r read, message=FALSE, warning=FALSE}
performance <- read_csv("data-in/2018-2019_936_ks2final.csv") %>% 
  clean_names()

head(performance)
```

There are over 300 columns of data in this file. For the purposes of this module we can condense this down to a more manageable number of columns.
```{r select}
performance <- performance %>% 
  select(urn,
         schname,
         nftype,
         reldenom,
         totpups,
         matprog,
         writprog,
         readprog)

head(performance)
```

Looking at the data types that R has determined, some columns have been read in as character types when they should be numeric. We can fix this for the `totpups`, `matprog`, `writprog` and `readprog` columns.
```{r data types, message=FALSE, warning=FALSE}
performance <- performance %>% 
  mutate(across(totpups:readprog, as.numeric))

head(performance)
```

# What is data?
Data is information and comes in different types:

* **Qualitative** data describes qualities. It is descriptive. There are two types of qualitative data:
    + **Ordinal** tells you about ordering (e.g. it ranks 8 out of 10)
    + **Nominal** simply describes or categorises something (e.g. it's pink)
* **Quantitative** data talks about quantities or numbers (e.g. 15cm high). There are two types of quantitative data:
    + **Discrete** can only take a set of possible values (e.g. £1.90)
    + **Continuous** can take on any value in a range (e.g. 12, 12.31. 12.34562)

Usually we are interested in data that comes from recording some quantity or quality of interest. We may know in advance what the possible values of the data could be, but until we collect the data we do not know the actual values. For example, we know that if we roll a dice we will get a number between 1 and 6, but we don't know which one until we roll it. We might know that the speed of a moving car at a particular location is between 0 and 100 miles per hour, but we don't know the exact speed until we measure it. When we want to measure a quantity or quality and it can take on a range of values like this, before we observe the actual value we refer to it as a **random variable**, or variable for short.

## What types of data can you find?
In our example data set we have the following fields and data types:

Column Name     Data Type               Description
--------------- ----------------------  -----------------
urn             Qualitative Nominal     Unique Reference Number of the school
schname         Qualitative Nominal     School name
nftype          Qualitative Nominal     School type
reldenom        Qualitative Nominal     Religious denomination
totpups         Quantitative Discrete   Total number of pupils
matprog         Quantitative Continuous Maths progress measure
writprog        Quantitative Continuous Writing progress measure
readprog        Quantitative Continuous Reading progress measure

Table: Fields and data types in the example data set

# What is statistics?
Statistics is a branch of maths that deals with analysing data and interpreting the results. Scientific research is often advanced by posing scientific questions and then answering these, and statistics answers those questions using data. For example, if you are looking at primary school performance, you might want to know if schools with a religious denomination achieve better maths results than those with no religious denomination. To answer this, you might first collect some data. For example, you could take a random sample of 100 schools; 50 of those have no religious denomination and 50 schools have a religious denomination. Then you would collect data on the maths performance of each school and use that data to determine whether having a religious denomination improves school performance in maths.

In order to answer the question, you would first look at the data, which might look something like this:

```{r maths performance}
# We'll create a new df called test_reldenom to hold our sample data for
# religious denomination and maths performance.
test_reldenom <- performance %>% 
  # First we'll select only the columns we need
  select(urn, reldenom, matprog) %>% 
  # Next we'll filter out any NA values in matprog
  filter(!is.na(matprog)) %>% 
  # Now let's create a new binary variable which determines if the school has
  # a religious denomination or not
  mutate(is_religious = if_else(
    reldenom %in% c("None", "Does not apply"),
    FALSE,
    TRUE)) %>% 
  # Now we'll subset the df so that we have two equal groups of 50 schools,
  # comparing religious against non-religious schools
  group_by(is_religious) %>% 
  slice_sample(n = 50) %>% 
  ungroup() %>% 
  # We can remove the reldenom column and re-order the remaining columns
  select(urn, is_religious, matprog)

test_reldenom
```

The table shows 100 schools, and the corresponding maths progress score for each. You can then divide them into 2 groups and create a chart with one plot for the group with a religious denomination and one plot for the group without a religious denomination, as shown below.
```{r boxplot}
ggplot(test_reldenom, aes(is_religious, matprog)) +
  geom_boxplot()
```

Looking at this data might tell us that having a religious denomination has possibly increased the maths progress score, but we would need to do a formal statistical analysis, which will take into account the different amounts of variability, to work this out for certain.

## The statistics cycle
Often we start out with an initial question that we want to answer with data. We collect the data and analyse it to answer that question, but in doing so that uncovers extra questions that we may want to go further and investigate. This process of analysis, refinement and reanalysis is generally how scientific research develops. We call this diagram the statistics cycle as shown below.